---
title: "An Introduction to Extreme Value Theory in R"
author: "Kate Saunders"
format: 
  revealjs:
    css: mathstyle.css
editor: visual
execute: 
  enabled: true
# server: shiny
bibliography: references.bib
---

## Motivation

<br> Estimating the probability of extreme events is useful for:\
<br>

-   Knowing how high to build a dam to reduce flood risk\
    <br>

-   Appropriately pricing insurance at given a location\
    <br>

-   Advising government about the risks from climate extremes <br>

## Today

-   Statistics of Extremes
-   Getting climate data in R
-   Learn how to fit extreme value distributions.\
    (diagnostics, model selection etc.)
-   Learn about return periods
-   Troubleshooting

# Background stats

## Observations vs Random Variables

::: incremental
[**Observations**]{.text-color} are things we've seen. <br>

-   Example: rolled a six or yesterday's temperature\
-   Use lower case letters to write observations $x$
:::

::: incremental
<br> [**Random variables**]{.text-color} are things we've yet to see

-   Example: next roll of the dice or tomorrow's temperature.\
-   Statistically: Use upper case letters to write random variables $X$.
:::

## Discrete Random Variables

::: incremental
Let $X$ be the random variable for a six sided dice roll then: <br>

-   Probability my next roll is a 4 is $\mathbb{P}(X = 4) = 1/6$

-   Probability my next roll is less than a 4 is $\mathbb{P}(X < 4) = 3/6$

-   The **probability mass function** for discrete random variables is $\mathbb{P}(X = x)$.

-   The **cumulative distribution function** for discrete random variables $\mathbb{P}(X \leq x)$
:::

## Discrete Random Variables ctd.

```{r, warnings = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)
```

```{r echo = FALSE}
dice_data = data.frame(x = 1:6, y = rep(1/6, 6)) |>
  mutate(cumsum_y = cumsum(y))

mass_plot <- ggplot(data = dice_data) + 
  geom_col(aes(x = x, y = y)) + 
  theme_bw() + 
  xlab("Dice Roll") + 
  ylab("P(X = x)") +
  ggtitle("Probability Mass Function") + 
  ylim(c(0,1))

disc_distribution_plot <- ggplot(data = dice_data) + 
  geom_col(aes(x = x, y = cumsum_y)) +
  theme_bw() + 
 xlab("Dice Roll") + 
 ylab("Pr(X <= x)") +
 ggtitle("Cumulative Distribution Function ") + 
 ylim(c(0,1))


mass_plot + disc_distribution_plot
```

## Continuous random variables

```{r echo = FALSE}
temp_values = seq(0, 45, length.out = 1000)
temp_mean = 20
temp_sd = 3.5

temp_density <- dnorm(temp_values, mean = temp_mean, sd = temp_sd) 
temp_distribution <- pnorm(temp_values, mean = temp_mean, sd = temp_sd)

temp_data_frame = data.frame(temp_values, temp_density, temp_distribution)

density_plot <- ggplot(data= temp_data_frame) + 
  geom_line(aes(x = temp_values, y = temp_density)) + 
  theme_bw() + 
  xlab("Temperature") + 
  ylab("f(x)") +
  ggtitle("Probability Density Function f(X = x)")

distribution_plot <- ggplot(data= temp_data_frame) + 
  geom_line(aes(x = temp_values, y = temp_distribution)) +
  theme_bw() + 
  xlab("Temperature") + 
  ylab("F(x)") +
  ggtitle("Probability Distribution Function F(X < x)")

density_plot + distribution_plot
```

## Law of Large Numbers

```{r echo = FALSE}
set.seed(1)
num_samples = seq(10,1000, by = 5)

fun_sample_mean <- function(i){mean(rnorm(i, mean = temp_mean, sd = temp_sd))}

sample_mean = sapply(num_samples, fun_sample_mean)
                 
ggplot(data = NULL) + 
  geom_point(aes(x = num_samples, y = sample_mean)) + 
  geom_hline(aes(yintercept = temp_mean), col = "red", linetype = "dotted") +
  theme_bw() + 
  xlab("Number of Samples") + 
  ylab("Sample Mean") + 
  ggtitle("Estimating the Mean")
```

<!-- ## Estimating the standard deviation -->

<!-- ```{r} -->

<!-- set.seed(1) -->

<!-- num_samples = seq(10,500, by = 10) -->

<!-- fun_sample_sd <- function(i){sd(rnorm(i, mean = temp_mean, sd = temp_sd))/sqrt(i-1)} -->

<!-- sample_sd = sapply(num_samples, fun_sample_sd) -->

<!-- ggplot(data = NULL) +  -->

<!--   geom_point(aes(x = num_samples, y = sample_sd)) +  -->

<!--   theme_bw() +  -->

<!--   xlab("Number of Samples") +  -->

<!--   ylab("Estimated Standard Deviation")  -->

<!-- ``` -->

## Central Limit Theorem

::: incremental
-   [**What is it?**]{.text-color}For large enough sample, the distribution of the sample mean will be approximately normal.

-   [**What do we need for this to be true:**]{.text-color} Continuous random variable with a finite mean and finite variance.

-   [**Why is that cool?**]{.text-color} No matter what the original data looks like, if you take the average of enough samples, the distribution of those averages will look like a normal distribution.

-   [**Why is that useful?**]{.text-color} Good for statistical inference, e.g. hypothesis testing, and estimating confidence intervals.
:::

# Extreme value theory

## Set up

Consider a continuous random variable $X$

<br><br>

with distribution function $F(x) = \mathbb{P}(X < x)$.

<br><br>

For a sequence of independent and identically distributed random variables, $X_1$, $X_2$, $\dots$, $X_n$ the maximum is

$$
M_{n} = \max \lbrace X_1, X_2, \dots X_{n} \rbrace.
$$

## Set up

The probability $M_n$ is less than some high quantile $z$ is

$$ \mathbb{P}(M_{n} < z) = \mathbb{P}(X_1 < z, X_2 < z, \dots , X_{n} < z)$$

<br> Using that the random variables are independent

$$ \mathbb{P}(M_{n} < z) = \mathbb{P}(X_1 < z) \times \mathbb{P}(X_2 < z), \times \dots \times \mathbb{P}(X_{n} < z)$$ <br> Using that the random variables are identically distributed

$$ \mathbb{P}(M_{n} < z) = \mathbb{P}(X < z)^n = \left[ F(z) \right]^n $$

## Set Up

But we don't know $F(z)$

We can try to estimate $F(z)$ from the data.

e.g For $m$ observations:

$$ F(z) = \mathbb{P}(X < z) = \dfrac{1}{m}\sum_{i = 1}^m \left( x_i < z \right). $$

That won't work well for extreme values of $z$.

## Set Up

<br> Other problem for large $n$ values, small errors in the estimation of $F(z)$ become big errors when we estimate $\left[F(z)\right]^n$. <br><br>

So how else can we estimate $\mathbb{P}(M_n < z)$?

## Limiting Distributions

While we don't know $\left[F(z)\right]^n$\
<br>

We do know what the limiting distribution is - call that $G(z)$!\
<br>

We can estimate $G(z)$ using extreme value theory .\
<br>

So for large enough $n$ provided the limit exists and is non-dengerate:

$$ \mathbb{P}(M_n < z) = \left[F(z)\right]^n \approx \lim_{n \rightarrow \infty} \left[F(z)\right]^n \rightarrow G(z; \mu, \sigma, \xi)$$

<!-- ## Extreme Value Theorem -->

<!-- A "Central Limit Theorem" but for extremes. -->

<!-- <span class="text-color">**What is it?**</span> For large enough sample, the distribution of the maxima will be an extreme value distribution. -->

<!-- <span class="text-color">**What do we need for this to be true:**</span> Continuous random variable, where samples to be identically and independently distributed -->

<!-- <span class="text-color">**Why is that cool?**</span> No matter what the original data looks like, if you take the sample maxima, the distribution of those maxima will look like a generalised extreme value distribution. -->

<!--  \<span class="text-color"\> \*\*Why is that useful?\*\*\</span\>  -->

<!-- Good for making statistical inferences about populations, hypothesis testing, and estimating confidence intervals AND we can now extrapolate beyond our observed data. -->

## Three different types

Depending on the data you start with you can end up with one of three non-degenerate limiting distributions: <br>

::: incremental
-   [I: Weibull]{.text-color} (Bounded extremes) <br>*e.g. Temperature*

-   [II: Gumbel]{.text-color} (Thin-tailed extremes) <br> *e.g.Storm Surge*

-   [III: Fréchet]{.text-color} (Heavy-tailed extremes) <br> *e.g Rainfall*
:::

Don't need to know which limiting regime before you start!

## GEV Distribution

Generalised Extreme Value (GEV) Distribution [@coles2001]:

$$  G(z; \mu, \sigma, \xi) = \exp \Bigg\lbrace{ -\left[1 + \xi \left( \dfrac{z - \mu}{\sigma}\right)  \right]^{-1 / \xi}\Bigg\rbrace} $$

Where:

-   $-\infty < \mu < \infty$ is the location parameter,\
-   $\sigma > 0$ is the scale parameter,\
-   $-\infty < \xi < \infty$ is the shape parameter,\
-   and $1 + \xi \left( \frac{z - \mu}{\sigma} \right) > 0$

## GEV Distribution in Shiny

Shiny app time

```{r eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE}
library(devtools)
devtools::install_github("ropensci/plotly")
library(extRemes)
library(tidyverse)
library(shiny)
```

```{r eval = FALSE}

#| context: server

# Define UI for application that draws a histogram
ui <- fluidPage(
  
  # Sidebar with a slider input for number of bins 
  sidebarLayout(
    
    sidebarPanel(
      sliderInput("loc",
                  "Location:",
                  min = -5,
                  max = 5,
                  value = 0,
                  step = 0.025,
                  animate = animationOptions(interval = 10)),
      
      sliderInput("scale",
                  "Scale:",
                  min = 0.5,
                  max = 3,
                  value = 1,
                  step = 0.025,
                  animate = animationOptions(interval = 10)),
      
      sliderInput("shape",
                  "Shape:",
                  min = -0.5,
                  max = 1,
                  value = 0,
                  step = 0.01)
    ),
    
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("gevPlot")
    )
  )
)

# Define server logic required to draw a histogram
server <- function(input, output) {
  
  output$gevPlot <- renderPlot({
    
    x = seq(-10, 20, length.out = 500)
    y_orig = devd(x, 0, 1, 0)
    y = devd(x, input$loc, input$scale, input$shape)
    y_p = pevd(x, input$loc, input$scale, input$shape)
    
    df = data.frame(x, y, y_orig, y_p)
    
    gev_plot <- ggplot(data = df) + 
      geom_line(aes(x=x, y=y_orig), linetype = "dotted") + 
      geom_line(aes(x=x, y=y), col = "blue") + 
      scale_x_continuous(limits = c(-10, 20)) +
      scale_y_continuous(limits = c(0, 0.6)) + 
      xlab("x") +
      ylab("y") +
      theme_bw()
    
    # ggplotly(gev_plot)
    gev_plot
    
  })
}

# Run the application 
shinyApp(ui = ui, server = server)


```

## Real world

[**Problem:**]{.text-color}<br> Climate variables like rainfall and temperature are not identically distributed.

<br>

::: incremental
That's okay. We have options.

-   Stratify your data\
    *e.g. fit to a stationary period, fit to Summer.*

-   Alternatively you can allow your GEV parameters to vary smoothly as a function of covariates.\
    *e.g. let* $\mu$ *vary with the year, vary with climate change*
:::

## Real world

[**Problem:**]{.text-color}<br> Climate variables like rainfall and temperature are not independent. The weather today, has an influence on the weather tomorrow.

<br>

::: incremental
That's okay.

-   We only worry if there is long-memory in the process.

-   The temperature today is independent of the temperature this time next year
:::

# Getting Started in R

## Packages

Let's install the packages we need for today:

```{r packages, echo = TRUE}
if(!require('tidyverse', character.only = TRUE))
  install.packages('tidyverse')

if(!require('rnoaa', character.only = TRUE)){
  install.packages("devtools")
  devtools::install_github("ropensci/rnoaa")
}

if(!require('extRemes', character.only = TRUE))
  install.packages('extRemes')
```

<br> Let's load those packages into our library:

```{r library, echo = TRUE}
library(tidyverse)
library(rnoaa)
library(extRemes)
```

## Getting Data

Using the `rnoaa` R package [@rnoaa], we can:

-   Access the daily observations in the Global Historical Climate Network (GHCN)

-   For Australia, these stations are the same as those monitored by the Bureau of Meteorology

-   Australian stations have IDs starting with "ASN"

```{r rnoaa, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}

# Get meta data
meta_data <- rnoaa::ghcnd_stations() 

# Restricts to Australian meta data
aus_code = "ASN"
aus_meta_data <- meta_data |>
    dplyr::filter(str_detect(id, aus_code))
```

## Search for a station

You are welcome to pick a station close to your home or your heart. <br>

**Conditions:** Pick something with a long record (\>50 years) that has elements of both "TMAX" and "PRCP".

```{r locnmatch, echo = TRUE}
# Search for stations within a radius of a location

lat_coord = -27.55; long_coord = 152.34

local_meta_data <- aus_meta_data |> 
  meteo_distance(lat = lat_coord, long = long_coord, radius = 5)

```

## Search for a station

May also like to look for a station by name

```{r namematch, echo = TRUE, eval = FALSE}
# Search for stations that partially match a name

search_term = "UNIVERSITY"

local_meta_data <- aus_meta_data |> 
  dplyr::filter(stringr::str_detect(name, search_term))
```

## Meta Data

```{r print}
local_meta_data
```

## Download the Data

Download the data using `meteo_pull_monitors`

Example using the station ID

```{r download1, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}
stn_id = "ASN00040082"
stn_data <- rnoaa::meteo_pull_monitors(monitors = stn_id,
                                    keep_flags =  TRUE,
                                    var = "all")
```

or using the station name

```{r download, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE}
stn_name = "UNIVERSITY OF QUEENSLAND GATTO"
stn_id <- local_meta_data |>
  dplyr::filter(name == stn_name) |>
  dplyr::select(id) |>
  dplyr::distinct() |>
  dplyr::pull()

# Downlaod the station ID
stn_data <- rnoaa::meteo_pull_monitors(monitors = stn_id,
                                    keep_flags =  TRUE,
                                    var = "all")
```

## Preprocess the Data

Light touch: Just doing the basics to tidy and format the data.

```{r preprocessing, echo = TRUE}
stn_data_preprocessed <- stn_data |>
  
  # Ensure Year format is a date 
  dplyr::mutate(year = lubridate::year(date)) |> 
  
  # Select the relevant columns
  dplyr::select(tmax, tmin, prcp, date, year) |> 
  
  # Pivot to a tidy long format
  tidyr::pivot_longer(cols = c(tmax, tmin, prcp), names_to = "element", values_to = "observation") |>
  
  # Change the unit of measurement
  dplyr::mutate(observation = as.numeric(observation)/10) |>
  
  # Filter out NA values (temperature)
  dplyr::mutate(observation = if_else(
    observation == 99 & element %in% c('tmax','tmin'),  
    NA, observation))
```

## Get the maxima

We want to get the highest observation each year. <br>\
Light touch: Restrict to years with more than 90% of the observations each year.

```{r blockmaxima, echo = TRUE}
stn_max <- stn_data_preprocessed |> 
  dplyr::group_by(year, element) |>
  dplyr::summarise(max = max(observation, na.rm = TRUE),
                   count = sum(as.numeric(!is.na(observation)))) |>
  ungroup() |>
  dplyr::filter(count > 365*0.9) |>
  select(-count) 
```

## Get the maxima

```{r}
stn_max
```

## Plot the Maxima

```{r maxplot, echo = FALSE, fig.width = 14, fig.height = 7, fig.align = 'center'}
maxima_plot <- ggplot(stn_max, aes(x = max)) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density(color = "red", linetype = "dashed", size = 1.5) +
  facet_wrap(~element, scale = "free") +
  theme_bw() + 
  xlab("Yearly Max") +
  ylab("Density") 

maxima_plot
```

-   Wettest Day of the year = Rx1d
-   Hottest Day of the year for Maximum Temp = Txx
-   Hottest Day of the year for Minimum Temp = Tnx

## Plot the Maxima

Code for producing the plot:

```{r maxplotcode, echo = TRUE, fig.align='center'}
ggplot(stn_max, aes(x = max)) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density(color = "red", linetype = "dashed", size = 1.5) +
  facet_wrap(~element, scale = "free") +
  theme_bw() + 
  xlab("Yearly Max") +
  ylab("Density") 
```

## Tidy the maxima

-   Transform the data to a wide format the data for fitting

-   Update to use the correct labels

```{r tidymax, echo = TRUE}

# Extract the block maxima
max_df = stn_max |> 
  tidyr::pivot_wider(names_from = element, values_from = max) |>
  as.data.frame() |>
  rename(rx1d = prcp, txx = tmax, tnx = tmin)

max_df
```

## Your turn

::: callout-tip
## Practice Time

-   Take some time now to explore the basic functions in `rnoaa`

-   Either use my code to download some station data

-   Or search for a station of interest to you and download that data

-   Once you have your data, create a plot to check your maxima

-   Also format your data to be ready for model fitting
:::

## Backups

If its helpful you can also directly read in the files I've just used

```{r eval = FALSE, echo = TRUE}
# Be sure to set your working directory first

aus_meta_data = read_csv("data/aus_meta_data.csv")

stn_data = read_csv("data/UoQGatton_data.csv")

```

# Fitting a GEV

Covering the basics

## Fitting

-   Practice using the precipitation maxima

-   Fit a stationary GEV, so constant parameters

-   $\mu = \mu_0$, $\sigma = \sigma_0$ and $\xi = \xi_0$.

-   Using the R package extRemes [@extRemes-2]

*Option 1:* Pass our data in as a vector

```{r vecfit, echo = TRUE}
# Extract the block maxima
rx1d = max_df$rx1d

# Fit the model
rx1d_gev_fit_vec = extRemes::fevd(x = rx1d, type = "GEV")

# Look at the parameters
rx1d_gev_fit_vec$results$par |> round(2)
```

<br> Sanity check: Do the parameters look reasonable?

## Fitting

*Option 2:* Pass our data in as a data frame

```{r dffit, echo = TRUE}
fitting_data = max_df |> dplyr::select(year, rx1d) 

# Fit the model
rx1d_gev_fit = fevd(x = rx1d, data = fitting_data, type = "GEV")

# Look at the parameters
rx1d_gev_fit$results$par |> round(2)
```

<br> Sanity check: Do the parameters look reasonable?

<br> Note you don't need to preselect your columns, I chose to to this to avoid dealing with NAs more directly.

## Model fitting summary

```{r fitsummary, echo = TRUE}
rx1d_gev_fit
```

## Diagnostic plot

```{r, echo = TRUE, fig.width = 8.5, fig.height = 6, fig.align = 'center'}
plot(rx1d_gev_fit)
```

## Return periods and return levels

We want to know the level that is exceeded on average 1 in every $100$ years:

$$\mathbb{P}(M_n < z_{100}) \approx G(z_{100}) = \dfrac{1}{100}. $$

Here we need to solve for $z_{100}$. This is what is known as the 1 in $100$ year return level.

<br> *In generality:* The return level $z_{p}$ associated with a return period $p$ is: $$\mathbb{P}(M_n < z_{p}) \approx G(z_{p}) = \dfrac{1}{p}. $$

## Return periods and return levels

*Approach 1:* Can use the quantile function

```{r qevd, echo = TRUE}
pars = rx1d_gev_fit$results$par
rl100 = qevd(p = 1/100, loc = pars[1], scale = pars[2], shape = pars[3], lower.tail = FALSE) 
```

A 1 in $100$ year event at this station is `r round(rl100)` mm. <br>

But ... <br>

That doesn't give CIs.

## Return periods and return levels

*Approach 2:* There is an inbuilt to estimate return levels from your fitted model with CIs

```{r eval = TRUE, echo = TRUE}
return.level(rx1d_gev_fit, return.period = 100, do.ci = TRUE)
```

## Annual Exceedance Probabilities

"1 in 100s" can be confusing.

::: incremental
-   It isn't an event that happens exactly once ever 100 years.

-   It doesn't mean we can't have two 1 in 100 year events in a row.

-   It makes no sense in a non-stationary climate

-   Now say the event with an Annual Exceedance Probability (AEP) of 0.01 instead
:::

## Return periods and return levels

Estimating a multiple return levels using your model:

```{r eval = TRUE, echo = TRUE}
return_periods = c(2:500)

return_levels = return.level(rx1d_gev_fit, return.period = return_periods, do.ci = TRUE)[,1:3] |> as.data.frame()

names(return_levels) = c("lower", "estimate", "upper")
return_levels$return_periods = return_periods
```

## Plot return periods and return levels

```{r returnplotcode, echo = TRUE}
# Get the 1 in 100 year return level
rl100 = return_levels |>
  dplyr::filter(return_periods == 100) |>
  dplyr::pull(estimate)


# Plot the return periods
return_period_plot <- ggplot() + 
  geom_ribbon(data = return_levels, aes(x = return_periods, ymin = lower, ymax = upper), alpha = 0.5, fill = "blue") + 
  geom_line(data = return_levels, aes(x = return_periods, y = estimate)) + 
  geom_line(data = NULL, aes(x = c(2, 100, 100), y = c(rl100, rl100,0)), linetype = "dashed") +
  scale_x_log10(breaks = c(2,5,10,20,50,100,250,500)) +
  theme_bw() + 
  xlab('Return Period') + 
  ylab('Return Level (mm)') + 
  ggtitle(paste("Extreme Rainfall at", stn_id))

```

## Plot return periods and return levels

```{r returnplot}
return_period_plot 
```

## Your turn

<br>

::: callout-tip
## Practice Time

-   Fit a stationary GEV to your data

-   Start with Rx1d (Rainfall)

-   Look at your parameters and diagnostic plots

-   Produce a plot of the return level

-   Extra: Add your data points to the plot (use empirical quantiles).

-   Skip ahead: Try fitting other variables

-   Extension: Recreate some of the other diagnostic plots
:::

```{r echo = FALSE}
emp_quantiles = ecdf(rx1d)(rx1d)
```

# Fitting a GEV

Non-stationarity and Model Selection

## Block Maxima data (Tmax)

Prepare our data:

-   Want a variable for year that is more intuitive

```{r echo = TRUE, eval = TRUE}
fitting_data = max_df |> 
  dplyr::select(year, txx) |>
  dplyr::filter(!is.na(txx)) |>
  dplyr::mutate(year_transform = year - 2024)

fitting_data
```

## Stationary GEV

**Stationary Model:**<br> $\mu = \mu_0$, $\sigma = \sigma_0$ and $\xi = \xi_0$.

```{r echo = TRUE, eval = TRUE}
txx_gev_fit0 = fevd(x = txx, data = fitting_data, type = "GEV")
txx_gev_fit0$results$par |> round(2)
```

## Non-stationary GEV

<br> **Non-stationary model:** Linear shift in $\mu$. <br> $\mu = \mu_0 + \mu_1 (\text{year} - 2024)$ <br> $\sigma = \sigma_0$ <br> $xi = \xi_0$

<br> **Code:** `location = ~year_transform`

```{r echo = TRUE}
txx_gev_fit1 = fevd(x = txx, data = fitting_data, type = "GEV", location = ~ year_transform)
txx_gev_fit1$results$par |> round(2)
```

## Fitting a GEV ctd.

**Non-stationary model:** Linear shift in location and scale: <br> $\mu = \mu_0 + \mu_1(\text{year} - 2024)$ <br> $\sigma = \sigma_0 + \sigma_1(\text{year} - \text{start year})$ <br> $\xi = \xi_0$

<br> **Code:** <br> `location = ~year_transform` <br> `scale = ~year_transform`

```{r echo = TRUE}
txx_gev_fit2 = fevd(x = txx, data = fitting_data, type = "GEV", location = ~year_transform, scale = ~year_transform)
txx_gev_fit2$results$par |> round(7)
```

<!-- ## Fitting a GEV ctd. -->

<!-- Linear shift in location and step jump in scale:  -->

<!-- <br> -->

<!-- <br> **Non-stationary model:** <br>  -->

<!-- $\mu = \mu_0 + \mu_1(\text{year} - 2024)$ <br>  -->

<!-- $\sigma = \sigma_0 + \sigma_1 \cdot \mathbb{I}(\text{year} > 1960 )$ <br> -->

<!-- $\xi = \xi_0$ -->

<!-- ```{r echo = TRUE} -->

<!-- fitting_data = fitting_data |> mutate(late = year - 1960) -->

<!-- txx_gev_fit3 = fevd(x = txx, data = fitting_data, type = "GEV", location = ~year_transform, scale = ~ late) -->

<!-- txx_gev_fit3$results$par |> round(2) -->

<!-- ``` -->

## Model Selection

Which model is better?

For **nested models** best to use the log-likelihood ratio test.

Nested means the parameter function for the larger model can be written as subset of the smaller model.

For non-nested models use an Information Criteria AIC/BIC - smaller is better.

## Model Selection

Model with a linear shift is best

```{r echo = TRUE, eval = TRUE}
#Stationary vs Linear shift in mu 
lr.test(txx_gev_fit0, txx_gev_fit1)[4]

#Stationary vs Linear shift in mu and in sigma 
lr.test(txx_gev_fit0, txx_gev_fit2)[4]

#Linear shift in mu vs Linear shift in mu and in sigma
lr.test(txx_gev_fit1, txx_gev_fit2)[4]

```

## Check diagnostics

Remember to look at the model you pick!

```{r}
plot(txx_gev_fit1)
```

## Your turn

<br>

::: callout-tip
## Practice Time

-   Fit a non-stationary GEV to your data

-   Start with txx (Maximum Temp)

-   Look at your parameters and diagnostic plots

-   Determine if climate change has a significant linear influence on extremes

-   You may also like to fit a GEV to tnx or rx1d

-   Extension: Try other functional forms to smoothly vary your covariates, suggestions include an indicator function or ENSO covariate.
:::

# Return Levels Under Non-stationartiy

## Conditional return revels

Can estimate the return level conditional on a given year.

```{r echo = TRUE}
return_periods = c(2:1000)

pars = txx_gev_fit1$results$par

mu0 = pars[1] 
mu1 = pars[2]
scale = pars[3]
shape = pars[4]

get_cond_mu <- function(conditional_year, mu0, mu1){
  mu = mu0 + mu1*(conditional_year - 2024)
  return(mu)
} 


rls_cond2024 = qevd(1/return_periods, loc = get_cond_mu(2024, mu0, mu1), scale = scale, shape = shape, lower.tail = FALSE)

rls_cond1924 = qevd(1/return_periods, loc = get_cond_mu(1924, mu0, mu1), scale = scale, shape = shape, lower.tail = FALSE)

cond_rp = data.frame(return_periods, rls_cond2024, rls_cond1924) |> 
  tidyr::pivot_longer(cols = 2:3, names_to = "cond", values_to = "rl")
```

## Plotting conditional return periods

```{r, echo = TRUE, fig.align = 'center'}
cond_plot <- ggplot(cond_rp) +
  geom_line(aes(x = return_periods, y = rl, group = cond, col = cond)) + 
  scale_x_log10(breaks = c(2,5,10,20,50,100,250,500)) +
  theme_bw() + 
  ylab("Return Level") + 
  xlab("Return Period") +
  scale_color_manual(labels = c("1924", "2024"), values = c("rls_cond2024" = "red", "rls_cond1924" = "blue")) +  xlim(c(0,150))

cond_plot
```

## Plotting conditional return periods

```{r fig.align='center'}
cond_plot
```

<!-- ## Data points  -->

<!-- ```{r} -->

<!-- fitting_data$mu = get_cond_mu(fitting_data$year, mu0, mu1) -->

<!-- fitting_data$cprob = NA -->

<!-- for(i in 1:nrow(fitting_data)){ -->

<!--   fitting_data$cprob[i] = pevd(fitting_data$txx[i], fitting_data$mu[i], scale, shape) -->

<!-- } -->

<!-- rl_1924 = qevd(fitting_data$cprob, loc = get_cond_mu(1924, mu0, mu1), scale = scale, shape = shape) -->

<!-- rl_2024 = qevd(fitting_data$cprob, loc = get_cond_mu(2024, mu0, mu1), scale = scale, shape = shape) -->

<!-- <!-- ``` -->

<!-- ## Data transformations 1 -->

<!-- *Step 1:* transform the data to a standard marginal form, $Z \approx GEV(0,1,0)$ -->

<!-- <br> Use `trans` in built. -->

<!-- ```{r echo = TRUE} -->

<!-- gumbel_data = trans(txx_gev_fit1) -->

<!-- ``` -->

<!-- <br> Note $Z \approx GEV(1,1,1)$ is also a common transform for univariate GEV marginals. -->

<!-- ## Data transformations 2 -->

<!-- *Step 2:* Transform the data to the conditional distribution. -->

<!-- <br> -->

<!-- Use the `revtrans.evd` -->

<!-- <br> -->

<!-- For 2024: -->

<!-- ```{r echo = TRUE} -->

<!-- cond_year = 2024 -->

<!-- transformed_obs_2024 = revtrans.evd(gumbel_data, loc = get_cond_mu(cond_year, mu0, mu1), scale = scale, shape = shape) -->

<!-- ``` -->

<!-- ## Data transformations 3 -->

<!-- *Step 3:* Find the probability of those events from the fitted model. -->

<!-- ```{r echo = TRUE} -->

<!-- cprob_obs_2024 = pevd(transformed_obs_2024, loc = get_cond_mu(2024, mu0, mu1), scale = scale, shape = shape) -->

<!-- ``` -->

<!-- Now we have everything we need to add these points to the plot based conditional on the year. -->

<!-- ## Rinse and repeat -->

<!-- Repeat for 1924. -->

<!-- ```{r echo = TRUE} -->

<!-- cond_year = 1924   -->

<!-- transformed_obs_1924 = revtrans.evd(gumbel_data, loc = get_cond_mu(cond_year, mu0, mu1), scale = scale, shape = shape) -->

<!-- cprob_obs_1924 = pevd(transformed_obs_1924, loc = get_cond_mu(cond_year, mu0, mu1), scale = scale, shape = shape) -->

<!-- ``` -->

<!-- Combine into the data frame for plotting. -->

<!-- ```{r echo = TRUE} -->

<!-- fitting_data$yr2024 = transformed_obs_2024 -->

<!-- fitting_data$prob2024 = cprob_obs_2024 -->

<!-- fitting_data$yr1924 = transformed_obs_1924 -->

<!-- fitting_data$prob1924 = cprob_obs_1924 -->

<!-- ``` -->

<!-- ## Plot Conditional Return Levels -->

<!-- ```{r, fig.align='center'} -->

<!-- ggplot() + -->

<!--   geom_line(data = cond_rp, aes(x = return_periods, y = rl, group = cond, col = cond)) + -->

<!--   geom_point(data = fitting_data, aes(x = 1/(1 - cprob_obs_2024), y = transformed_obs_2024), col = "red") + -->

<!--    geom_point(data = fitting_data, aes(x = 1/(1 - cprob_obs_1924), y = transformed_obs_1924), col = "blue") + -->

<!--   scale_x_log10(breaks = c(2,5,10,20,50,100,250,500)) + -->

<!--   theme_bw() +  -->

<!--   ylab("Return Level") +  -->

<!--   xlab("Return Period") + -->

<!--   scale_color_manual(labels = c("1924", "2024"), values = c("rls_cond2024" = "red", "rls_cond1924" = "blue")) +  -->

<!--   xlim(c(0,150)) -->

<!-- ``` -->

## Your turn

<br><br>

::: callout-tip
## Practice Time

-   Produce a plot of the conditional return levels

-   Extension: Use bootstrapping to simulate data and estimate CIs for your conditional return levels
:::

# Tips and tricks

## Default

<br>

Default optimiser: Maximum Likelihood

```{r echo = TRUE}
set.seed(1)
timeA = Sys.time()
example_gev_fit_mle = fevd(x = revd(100), type = "GEV")
timeB= Sys.time()
timeB - timeA

example_gev_fit_mle$results$par |> round(2)
```

## LMoments is quick!

If you are fitting a lot of stationary GEVs you might like LMoments. <br>

This is because LMoments is a non-parametric estimator.

```{r echo = TRUE}
set.seed(1)
timeC = Sys.time()
example_gev_fit_lmoments = fevd(x = revd(100), type = "GEV", method = "Lmoments")
timeD= Sys.time()
timeD - timeC

example_gev_fit_lmoments$results |> round(2)
```

That's `r (as.numeric(timeB - timeA) / as.numeric(timeD - timeC)) |> round(2)` times as fast.

## Quick debugs - use.phi

-   If your model fails to fit one of the common reasons is that the scale parameter is negative.

-   You can force the optimiser to fit the log of the scale parameter instead to avoid this - set `use.phi = TRUE`

```{r usephi, echo = TRUE}
set.seed(1)
example_gev_fit_phi = fevd(x = revd(100), type = "GEV", use.phi = TRUE)

example_gev_fit_phi$results$par

exp(example_gev_fit_phi$results$par[2]) |> as.numeric() |> round(2)
```

## Optimiser is finicky

-   Still having problems with your model fit, try initialising your parameter estimates using `initial`

-   Lmoments is useful for automatically setting a starting point

```{r initial, echo = TRUE}
init_par = list("location" = -0.02, "scale" = 0.94, "shape" = 0.08)

example_gev_fit_init = fevd(x = revd(100), type = "GEV", initial = init_par)

example_gev_fit_init$results$par |> round(2)
```

-   Still not working, try a different starting point

-   You should also check your data here for any problem observations

## Handling a bad fit

Fit enough of these distributions you will come into troubles with the optimiser.

Example function replicating a failed fit:

```{r, echo = TRUE}
get_pars_from_model = function(vec){
  
  rand_samp = sample(c(0,0,0,0,1), 1)
  
  if(as.logical(rand_samp))
    stop("Too bad your fit randomly fails :(")
  
  model_fit = fevd(vec)
  pars = model_fit$results$par |> round(3)
  return(pars)
  
}

get_pars_from_model(revd(100))
```

## Trycatch

When running a script or fitting multiple distributions to avoid optimiser errors use `tryCatch()`.

```{r echo = TRUE}
try_fit <- function(vec){
  tryCatch(
    {
      pars <- get_pars_from_model(vec)
      return(c(pars, "good"))
    }, 
    warning = function(w){
      message(w)
      return(c(NA, NA, NA, "warning"))
    },
    error = function(e){
      message(e)
      return(c(NA, NA, NA, "error"))
    }
  )
}
try_fit(revd(100))
```

## Handling a Bad Fit

Here is an example:

```{r echo = TRUE}
```

```{r echo = TRUE}
set.seed(1)
results = data.frame(matrix(NA, 10,4))
for(i in 1:10){
  results[i, ] = try_fit(revd(100))
}

results
```

# Fitting a Generalised Pareto

Covering the basics

## So far

::: callout-tip
## GEVs

::: incremental
-   GEV distributions are fit to data that is the largest observation in a block\
    *eg. Wettest day of the year*

-   In a given year, there may be more than one extreme observation

-   In some years, maybe no observations are extreme

-   But what if you want to use all the extreme observations to fit a distribution\
    ie. $\mathbb{P}(X > y + u | X > u)$
:::
:::

## Generalised Pareto Distribution

::: callout-tip
## Can fit a Generalised Pareto Distribution (GP, PoT)

::: incremental
-   Closely related to the GEV

-   Model is fit to all observations exceeding a high threshold, $u$.

-   The threshold, $u$, needs to be extreme, so that the approximation to limiting distribution is reasonable. *e.g* $q_u >= 0.95$

-   Picking the threshold is a classic bias-variance trade-off.

-   Note the GP distribution has two parameters not three, $\tilde{\sigma}$ and $\xi$.

-   $\tilde{\sigma}$ is sort of like the scale parameter. $\tilde{\sigma} = \sigma + \xi(u - \mu)$

-   Note that $\tilde{\sigma}$ is a linear function of $u$ (can use this to set the threshold).

-   $\xi$ is the same as in the GEV.
:::
:::

## Quick example

```{r echo = TRUE}
# Get our precipitation observations 
prcp_obs = stn_data_preprocessed |>
  dplyr::filter(element == "prcp") |>
  dplyr::pull(observation)

# Estimate a high threshold
prcp_threshold = quantile(prcp_obs, 0.975, na.rm = TRUE)

# Filter your data to the exceedances   
prcp_exceedances = stn_data_preprocessed |>
  dplyr::filter(element == "prcp" & 
                  observation > prcp_threshold) |>
  dplyr::pull(observation)

# Fit a GP
prcp_gp_fit = fevd(prcp_exceedances, threshold = prcp_threshold, type = "GP")
  
# Parameters 
prcp_gp_fit$results$par
```

## Check your diagnostics

```{r}
plot(prcp_gp_fit)
```

## Other considerations

::: callout-tip
## Return Periods

::: incremental
-   For GEVs, the return periods were the number of *years* on average you expect to wait to see an event of a given return level

-   But for GEVs, each observation represented a year

-   For GPs, the return period is the the number of *observations* on average you expect before you see an event of a given level.

-   As for GPs, you can have more than one observation per year

-   In a stationary context, still quite interprettable, in a non-stationary context it gets a bit trickier.

-   For these reasons, many people prefer a GEV.
:::
:::

See Chapter 4 and 5 of [@coles2001].

## Your turn

<br><br>

::: callout-tip
## Practice Time

-   Take some time now to fit a Generalised Pareto

-   Extra: Look at how many extremes you have per year. (Can use the count code from earlier when we looked at how many observations per year.)

-   We will also be wrapping up today's tutorial shortly, so get to some help to wrap up your code projects.
:::

# Overview

## Summary

::: callout-tip
## Today:

-   Learnt about the basics of extreme value distributions

-   Covered how to fit these distributions in R

-   Looked at return periods and the nuance of fitting these

-   Covered some tips and tricks for fitting these distributions

-   And I hope we've had fun!
:::

## References

<!-- # Kate notes -->

<!-- Return level plot - add dots -->

<!-- Return level plot - non-stationary dots :( -->

<!-- Fix reference slide -->
